Containers and R---the Rockerverse and beyond

### 2019-11-05 manuscript.Rmd
update introduction

The R community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
All this cements the role of R as the lingua franca of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of Docker [CITATION?] as a general tool for development, distribution, and deployment of software.
Combining both these topics, the _Rocker Project_([https://www.rocker-project.org/](https://www.rocker-project.org/)) was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases as described in \citet{RJ-2017-065}.
The considerable uptake and continued evolution of the  _Rocker_ suite of containers has lead to numerous projects extending or building upon _Rocker_ images ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects.
It also introduces related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of containers in, with, and for `R`.
This diversity includes demonstrators, early prototypes, and mature projects.

@@

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing multiple concurrent applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability feature had been realized.
But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCES!].   
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker Project provides several lines of containeres spanning to from building blocks with R-release or R-devel to containers with RStudio~Server and Shiny~Server.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
[Maybe at a later point add forward reference to other topics building on Rocker described here.]

>>>>

The R community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
All this cements the role of R as the lingua franca of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of software.
Combining both these topics, the Rocker Project ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see details next Section).
The considerable uptake and continued evolution of Rocker images has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: demonstrators, prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

@@

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing multiple concurrent applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability feature had been realized.
But Docker improved on virtualization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This makes for more efficient use of system resources and allowed another order of magnitude in terms of scalability of deployment [REFERENCE!].   
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the Rocker Project has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with RStudio~Server and Shiny~Server, to domain-specific containers such as `geospatial`.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
[Maybe at a later point add forward reference to other topics building on Rocker described here.]
